\chapter{Introduction}

Video games are really complex products, but it is rather simple to point out three elements that mainly influence their commercial success: \<visuals>, \[gameplay]\footnote{The specific way in which players interact with the game.} and \<narrative>. If in the past visual improvements were impressively fast, with games looking more and more realistic from year to year, lately this progress consistently slowed down, leaving narrative and gameplay as the main selling points. Since video games are interactive products, the latter is the one that influences user experience the most. Gameplay is defined by a set of rules commonly referred as \<game design>. To the player, game design is presented in a tangible way via \<level design>, which consists in the creation of the worlds where the game takes place. This is a critical component, since an inadequate level design can easily compromise the whole experience.

\par

One of the most successful video game genre is the one of \[First Person Shooters], that thanks to its first player perspective allows the user to experiment a complete immersion in the game world. From the very beginning, this has required a close attention to level design, that underwent a constant evolution, up to the stable situation of the recent years. Furthermore, the ever-increasing success of competitive multiplayer added a new level of complexity to the creation of maps, that need to support different game modes, play styles and interactions, allowing a fun and challenging gameplay to arise naturally.

\section{Motivations and purpose}

Despite the importance of level design for the FPS genre, the video game industry has never attempted a scientific approach to this field. Consequently, game design is a rather abstract discipline, with no common vocabulary or well-defined standards, but rather based on the experience of who is working in this field from many years. This affects also the related literature, that is confined to listing the most used patterns and conventions, without focusing on why they work.

\par

In the last years, instead, academic environments started to address this discipline with increasing interest. The researches performed in this field revolve around the identification and definition of design patterns and design techniques, with a deep analysis of how and why they work, and to the creation of novel approaches to level design. Some of these techniques try to automate the design process by employing \<procedural generation>, often combined with \<evolutionary algorithms>.

\par

These methods have achieved interesting results, but they are often criticized for employing matches played by \<artificial agents> as a \<validation method>. Since the behavior of an AI, no matter how advanced, is way less complex than the one of a human player, this approach impairs the scientific value of the final results. Nevertheless, researchers are forced to use AI to fasten up their experiment, because organizing play-test sessions requites too much time and resources. 

\par

The aim of this thesis is to solve this problem, by introducing an open-source \<framework> capable of deploying online experiments to collect data from real users via a browser FPS game. Using this framework, we also explore the possibility of using \<Graph Theory> to populate procedural generated maps with spawn-points and various kinds of resources.

\section{Synopsis}

The contents of the thesis are the followings:

\par \mbox{}

In the second chapter we describe the state of the art of first person shooters, both in academic research and in commercial games, paying attention on how level design practices and procedural content generation are applied in this genre. We also list some examples of how Graph Theory is employed in video games.

\par
\mbox{}

In the third chapter we present the framework that we have developed, analyzing its features and its overall structure.

\par
\mbox{}

In the fourth chapter we present a tool that uses Graph Theory to place resources in procedurally generated levels, with an overview of the theory and the assumptions behind it.

\par
\mbox{}

In the fifth chapter we describe an experiment performed with our framework aimed at validating the framework itself and the tool described in the previous chapter.

\par 
\mbox{}

In the concluding chapter we evaluate the obtained results and we analyze the potential future developments of this work.